### 2022-12-29 paper.Rmd
editing and moved partition example back into paper

Anticlustering refers to the process of partitioning elements into disjoint groups with the goal of obtaining high between-group similarity and high within-group heterogeneity [@brusco2019; @spath1986; @valev1998]. Anticlustering thereby reverses the logic of its better known twin---cluster analysis---which seeks homogeneity within clusters and separation between clusters [@rokach2005; @steinley2006]. Anticlustering has many applications in research psychology [@steinley2006; @brusco2019; @papenberg2020]. Examples include splitting tests into parts of equal difficulty [@gierl2017], assigning students to work groups [@baker2002], and assigning stimuli to different, but parallel experimental conditions [@lahl2006]. Solving anticlustering problems "by hand" is a tedious and time-consuming task, and the quality of manual partitioning is usually subpar. Fortunately, anticlustering problems can be formalized as mathematical optimization problems [e.g., @baker2002; @brusco2019; @spath1986; @fernandez2013] and accessible open source software solutions to tackling these problems exist [@papenberg2020].

Anticlustering methods are characterized by (a) an objective function that quantifies between-group similarity and/or within-group heterogeneity, and (b) an algorithm that determines how elements are assigned to groups to maximize the objective function. Several anticlustering objective functions use pairwise dissimilarity ratings such as the Euclidean distance as input. The most prominent criterion of this kind is the *diversity*, which is the total sum of dissimilarities between elements within the same group [@brusco2019; @gallego2013]. By summing all pairwise dissimilarities within the same group, the diversity reflects the total degree of within-group heterogeneity. For anticlustering, high within-group heterogeneity is desired, which simultaneously ensures high between-group similarity.[^equalsizedgroups] Another important anticlustering criterion is the *dispersion*, which is the minimum dissimilarity between any two elements within the same group [@fernandez2013]. Maximizing the dispersion increases the within-group heterogeneity by ensuring that any two elements in the same group as dissimilar from each other as possible. @brusco2019 convincingly argued that anticlustering applications striving for within-group heterogeneity should incorporate both dispersion and diversity, and they presented a bicriterion algorithm to approximate the Pareto efficient set of solutions according to both criteria.

@@

Oftentimes, partitioning applications in psychology are focused on ensuring between-group similarity rather than within-group heterogeneity (even though both goals oftentimes coincide). High between-group similarity is for example desirable when preparing experimental conditions that use different stimulus sets, that should however be as similar as possible with regard to response-relevant attributes [@lahl2006]. Rather than relying on pairwise (dis)similarity ratings, stimulus selection tasks are usually conducted on the basis of attribute data. That is, individual stimuli are numerically coded on relevant dimensions [@dry2009]. For stimuli in psycholinguistic experiments, attributes may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attributes are binary and may represent the presence or absence of a feature [@tversky1977]. In the context of anticlustering, one approach for handling attribute data is to compute an appropriate dissimilarity measure such as the pairwise Euclidean or squared Euclidean distance across the set of attributes [@brusco2019]. Subsequently, a distance-based anticlustering criterion such as the diversity can be optimized to obtain a partitioning of the stimuli. A different anticlusting approach relies on clustering objectives that can directly be computed using the attribute values. K-means is probably the best-known objective of this kind. In k-means clustering, the sum of the squared Euclidean distances between data points and their cluster centers is minimized, usually using the k-means heuristic [@jain2010; @steinley2006; @brusco2006branch]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful approach when aiming for high between-group similarity [@spath1986; @valev1998; @valev1983]. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. However, other distribution characteristics---such as the variance---are not targeted. @papenberg2020 showed that k-means anticlustering tends to over-optimize similarity with regard to the mean; even an entirely random group assignment usually resulted in more similar variances. When requiring overall between-group similarity, the negligence of all distribution characteristics other than the mean is misguided, because similar means can be obtained even when the underlying distributions are clearly different [e.g., @anscombe1973].

@@

For the purpose of problem formalization, I adopt the notation @steinley2006 provided in his comprehensive synthesis of half a century of research on k-means. K-means anticlustering is used to partition $N$ elements each having $P$ attributes into $K$ groups $(C_1, \ldots, C_K)$, where $C_k$ is the set of $n_k$ objects in the $k$'th group. The partitioning is subject to the requirement that each element is assigned to exactly one group, formalized by the following two constraints [@papenberg2020]:

$$
\bigcup\limits_{k = 1}^{K} C_j = X
$$

$$
C_j \cap C_k = \emptyset, \; \forall j, k \in \{1, \ldots, K\}, \; j \ne k
$$

In anticlustering applications, an additional constraint is usually imposed on the group sizes $n_k$, where the most common restriction requires that all groups have equal size:

$$
n_k = \frac{N}{K}, \forall k \in \{1, \ldots, K\}
$$

@@

In the following, I present a variation of the $\mathit{SSE}$ that quantifies how the variance of numeric data differs between groups. To motivate the new criterion, we first consider a set of one-dimensional observations $(x_1, \ldots, x_N)$. The sample variance of a set of observations $x = (x_1, \ldots, x_N)$ is computed as the mean of the squared distances between observations and their mean $\overline{\mathbf{x}}$:

@@

Note that $\mathit{SSE_{Var}}$ has the same form as the standard $\mathit{SSE}$. However, the input data reflects the squared difference between each data point and the attribute mean instead of the raw data points.

@@

It is unlikely that an exclusive maximization of $\mathit{SSE_{Var}}$ (i.e., equalizing variances) is of interest without also considering the $\mathit{SSE}$ (i.e., equalizing means). A combination of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$, however, can be employed to simultaneously optimize both objectives, which is a reasonable idea when striving for overall between-group similarity. The most simple approach for such a bicriterion optimization is the weighted sum approach [@naidu2014; @marler2010]. That is, both criteria are computed independently from each other and are then combined into a single criterion through a weighted sum:

@@

@@

Figure 1 shows that with increasing $N$, k-means and the k-plus methods converged on the same level with regard to minimizing $\Delta \mathit{M}$. This is an important observation because it shows that including the k-plus term in the optimization process does not necessarily prevent minimizing discrepancy in means; instead, multiple objectives can be addressed at the same time---especially when $N$ increases. However, while k-plus anticlustering is capable of addressing multiple objectives simultaneously, increasing the number of optimization criteria apparently aggravates fulfilling each single criterion. Hence, k-plus skew/kurtosis and k-plus correlation performed worse at minimizing $\Delta \mathit{SD}$ than standard k-plus. Still, every k-plus method did what it was supposed to do: k-plus skew/kurtosis was best at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$, k-plus correlation was best at minimizing discrepancy with regard to $\Delta \mathit{Cor}$. At the same time, these two k-plus objectives maintained a comparably good level of addressing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$, outperforming diversity anticlustering.

It is maybe surprising that k-plus skew/kurtosis does not seem to have performed much better than diversity anticlustering with regard to skewness and---in particular---kurtosis, even though the method has been specifically tailored to these criteria. Figure 1 gives a more fine grained assessment of this observation by splitting the results by $N$: Surprisingly, for small $N$, k-plus skew/kurtosis was even worse than diversity anticlustering at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$. However, with increasing $N$, it clearly outperformed diversity anticlustering. These results illustrate the cost of optimizing several criteria at the same time: for low $N$, k-plus seems skew/kurtosis to be preoccupied with fulfilling the objectives $\Delta \mathit{M}$ and $\Delta \mathit{SD}$. Increasing $N$ then facilitated to also address $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$.

>>>>

Anticlustering refers to the process of partitioning elements into disjoint groups with the goal of obtaining high between-group similarity and high within-group heterogeneity [@brusco2019; @spath1986; @valev1998]. Anticlustering thereby reverses the logic of its better known twin---cluster analysis---which seeks homogeneity within clusters and separation between clusters [@rokach2005]. Anticlustering has many applications in research psychology [@steinley2006; @brusco2019; @papenberg2020]. Examples include splitting tests into parts of equal difficulty [@gierl2017], assigning students to work groups [@baker2002], and assigning stimuli to different, but parallel experimental conditions [@lahl2006]. Solving anticlustering problems "by hand" is a tedious and time-consuming task, and the quality of manual partitioning is usually subpar. Fortunately, anticlustering problems can be formalized as mathematical optimization problems [e.g., @baker2002; @brusco2019; @spath1986; @fernandez2013] and accessible open source software solutions to tackling these problems exist [@papenberg2020].

Anticlustering methods are characterized by (a) an objective function that quantifies between-group similarity and/or within-group heterogeneity, and (b) an algorithm that determines how elements are assigned to groups to maximize the objective function. Several anticlustering objective functions use pairwise dissimilarity ratings such as the Euclidean distance as input. The most prominent criterion of this kind is the *diversity*, which is the total sum of dissimilarities between elements within the same group [@brusco2019; @gallego2013]. By considering pairwise dissimilarities from between elements in the same group, the diversity reflects the total degree of within-group heterogeneity. High within-group diversity simultaneously ensures high between-group similarity.[^equalsizedgroups] Another important anticlustering criterion is the *dispersion*, which is the minimum dissimilarity between any two elements within the same group [@fernandez2013]. Maximizing the dispersion increases within-group heterogeneity by ensuring that any two elements in the same group as dissimilar from each other as possible. @brusco2019 convincingly argued that anticlustering applications striving for within-group heterogeneity should incorporate both dispersion and diversity, and they presented a bicriterion algorithm to approximate the Pareto efficient set of solutions according to both criteria.

@@

Oftentimes, partitioning applications in psychology are focused on between-group similarity rather than within-group heterogeneity (even though both goals oftentimes coincide). High between-group similarity is for example desirable when preparing experimental conditions that use different stimulus sets, that should however be as similar as possible with regard to response-relevant attributes [@lahl2006]. Rather than relying on pairwise (dis)similarity ratings, stimulus selection tasks are usually conducted on the basis of attribute data. That is, individual stimuli are numerically coded on relevant dimensions [@dry2009]. For stimuli in psycholinguistic experiments, attributes may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attributes are binary and may represent the presence or absence of a feature [@tversky1977]. In the context of anticlustering, one approach for handling attribute data is to compute an appropriate dissimilarity measure such as the pairwise Euclidean or squared Euclidean distance across the set of attributes [@brusco2019]. Subsequently, a distance-based anticlustering criterion such as the diversity can be optimized to obtain a partitioning of the stimuli. A different anticlusting approach relies on clustering objectives that can directly be computed on the attribute values. K-means is probably the best-known objective of this kind. In k-means clustering, the sum of the squared Euclidean distances between data points and their cluster centers is minimized, usually using the k-means heuristic [@jain2010; @steinley2006; @brusco2006branch]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful approach when aiming for high between-group similarity [@spath1986; @valev1998; @valev1983]. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. However, other distribution characteristics---such as the variance---are not targeted. @papenberg2020 showed that k-means anticlustering tends to over-optimize between-group similarity with regard to the mean at the cost of similarity in spread; even an entirely random group assignment usually resulted in more similar variances. When requiring overall between-group similarity, the negligence of distribution characteristics other than the mean is misguided, because similar means can be obtained even when the underlying distributions are clearly different [e.g., @anscombe1973].

@@

For the purpose of problem formalization, I adopt the notation @steinley2006 provided in his comprehensive synthesis of half a century of research on k-means. K-means anticlustering is used to partition $N$ elements each having $P$ attributes into $K$ groups $(C_1, \ldots, C_K)$, where $C_k$ is the set of $n_k$ objects in the $k$'th group. The partitioning is subject to the requirement that each element is assigned to exactly one group. In anticlustering applications, an additional constraint is usually imposed on the group sizes $n_k$, where the most common restriction requires that all groups have equal size [@papenberg2020].

@@

In the following, I present a variation of the $\mathit{SSE}$ that quantifies how the variance of numeric data differs between groups. To motivate the new criterion, we first consider a set of one-dimensional observations. The sample variance of a set of observations $x = (x_1, \ldots, x_N)$ is computed as the mean of the squared distances between observations and their mean $\overline{\mathbf{x}}$:

@@

Note that $\mathit{SSE_{Var}}$ has the same form as the standard $\mathit{SSE}$. However, the input data reflects the squared difference between each data point and the attribute mean instead of the raw data points. Maximizing $\mathit{SSE_{Var}}$ will lead to similar variances across the $K$ groups in each of the $P$ attributes.

@@

It is unlikely that an exclusive maximization of $\mathit{SSE_{Var}}$ (i.e., equalizing variances) is of interest without also considering the standard $\mathit{SSE}$ (i.e., equalizing means). A combination of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$, however, can be employed to simultaneously optimize both objectives, which is a reasonable idea when striving for overall between-group similarity. The most simple approach for such a bicriterion optimization is the weighted sum approach [@naidu2014; @marler2010]. That is, both criteria are computed independently from each other and are then combined into a single criterion through a weighted sum:

@@

# First rendering is slow because the data needs to be generated
path <- "partition_objectives.csv"

if (!file.exists(path)) {
  set.seed(2207) # 2207
  # Generate data and all partitions
  N <- 14
  K <- 2
  partitions <- generate_partitions(N, K)
  features <- rnorm(N, 15, 3)

  # Compute objectives for each partition
  # For each partition: Compute SSE objective, and SSE(VAR)
  # 1. SSE
  all_objectives_kMeans <- sapply(
    partitions,
    FUN = var_objective,
    features = features
  )

  squared_distances_from_mean <- squared_from_mean(as.matrix(features))

  # 2. SSE(VAR)
  all_objectives_kVar <- sapply(
    partitions,
    FUN = var_objective,
    features = squared_distances_from_mean
  )

  # 3. SSE(KPLUS)
  # Use standardization to compute k-plus criterion!
  all_objectives_kPlus <- sapply(
    partitions,
    FUN = var_objective,
    features = scale(cbind(features, squared_distances_from_mean))
  )

  df <- data.frame(
    kmeans = all_objectives_kMeans,
    kvar = all_objectives_kVar,
    kplus = all_objectives_kPlus
  )

  write.table(
    df, path, row.names = FALSE, sep = ","
  )
} else {
  df <- read.csv(path)
}

When optimizing a bicriterion objective such as $SSE_{kplus}$, the ideal outcome would be to find a partition where both the means and the variances are very similar between groups, i.e., a which partition where both the standard $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ are close to the global optimum value. Generally, research on multi-objective optimization has shown that cases are rare where a single solution satisfies all criteria to an optimal degree [@ferligoj1992direct]. Usually, a tradeoff is necessary. However, a motivating example illustrates that aiming to simultaneously optimize similarity of means and variances is not only desirable, but may actually be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatterplot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. Note that for both criteria, a larger value is better, i.e., indicates higher similarity with regard to means and variances, respectively. Therefore, to obtain overall between-group similarity, a partition should be chosen that is depicted in the upper right corner of the scatterplot. However, optimal k-means anticlustering would create two groups that are not particularly similar with regard to their variance. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (the triangular symbol in Figure 1) is not well-suited to optimize $\mathit{SSE}$ at the same time. The diamond symbol represents the partition that maximizes (the unweighted) $SSE_{kplus}$, i.e., the sum of $\mathit{SSE}$ and $\mathit{SSE_{Var}}$. This partition is located in the right upper corner and yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$.

```{r, fig.cap = "The relationship between the objectives $SSE$ and $SSE_{Var}$ across all ways to partition $N = 14$ normally distributed random values into $K = 2$ groups. The partition represented by the triangular minimizes the difference in variances between groups; the partition represented by the square minimizes the difference in means between groups; the partition represented by the diamond maximizes $SSE_{kplus}$ and represents a good tradeoff, satisfying both criteria to an almost optimal degree.", fig.height = 5, fig.width = 5}

# Plot SSE vs SSE(VAR)
plot(
  df[, c("kmeans", "kvar")], 
  pch = 4, col = "darkgrey", 
  las = 1, cex = .5,
  xlab = "SSE",
  ylab = "SSE_Var"
)

# Illustrate best partition wrt difference in means (i.e., SSE)
points(
  df$kmeans[which.max(df$kmeans)],
  df$kvar[which.max(df$kmeans)],
  cex = 1.2, pch = 22, bg = "white", lwd = 1.7
)

# Illustrate best partition wrt difference in variance (i.e., SSE(VAR))
points(
  df$kmeans[which.max(df$kvar)],
  df$kvar[which.max(df$kvar)],
  cex = 1.2, pch = 24, bg = "white", lwd = 1.7
)

# Illustrate best partition wrt k-plus criterion
points(
  df$kmeans[which.max(df$kplus)],
  df$kvar[which.max(df$kplus)],
  cex = 1.2, pch = 23, bg = "white", lwd = 1.7
)
```

@@

Figure 2 shows that with increasing $N$, k-means and the k-plus methods converged on the same level with regard to minimizing $\Delta \mathit{M}$. This is an important observation because it shows that including the k-plus term in the optimization process does not necessarily prevent minimizing discrepancy in means; instead, multiple objectives can be addressed at the same time---especially when $N$ increases. However, while k-plus anticlustering is capable of addressing multiple objectives simultaneously, increasing the number of optimization criteria apparently aggravates fulfilling each single criterion. Hence, k-plus skew/kurtosis and k-plus correlation performed worse at minimizing $\Delta \mathit{SD}$ than standard k-plus. Still, every k-plus method did what it was supposed to do: k-plus skew/kurtosis was best at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$, k-plus correlation was best at minimizing discrepancy with regard to $\Delta \mathit{Cor}$. At the same time, these two k-plus objectives maintained a comparably good level of addressing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$, outperforming diversity anticlustering.

It is maybe surprising that k-plus skew/kurtosis does not seem to have performed much better than diversity anticlustering with regard to skewness and---in particular---kurtosis, even though the method has been specifically tailored to these criteria. Figure 2 gives a more fine grained assessment of this observation by splitting the results by $N$: Surprisingly, for small $N$, k-plus skew/kurtosis was even worse than diversity anticlustering at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$. However, with increasing $N$, it clearly outperformed diversity anticlustering. These results illustrate the cost of optimizing several criteria at the same time: for low $N$, k-plus seems skew/kurtosis to be preoccupied with fulfilling the objectives $\Delta \mathit{M}$ and $\Delta \mathit{SD}$. Increasing $N$ then facilitated to also address $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$.
